Outline flowCore paper

### 2008-08-12 paper/manuscript.tex
couple of quick edits, will read on bus tomorrow in depth. Looking great! See email for suggestions for Figures.
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@33194 bc3139a8-67e5-0310-9ffc-ced21a209358

The notion of classes from an object-oriented programming language
provides one coherent way to describe these richer data structures,
and functions or methods that work on them allow for interaction and
manipulation. Many of the currently available software solutions offer
only limited support for such self-contained structures, or make use
of containers like XML or binary storage formats that are designed
specifically for particular user interfaces and hence are not easily
amenable to programmatical access. In addition, the closed-source
nature of these products often makes it impractical to integrate them
into analysis pipelines. In this manuscript we describe classes for
FCM data analysis and their implementation in R, however, they could
just as easily be implemented in any other language (eg Java,
C++). Software written in those languages could use similar data
structures, thereby simplifying communication and the interchange of
data between analysis tools.

The \Rpackage{flowCore} framework presented here can import and
process raw data FCS files along with their complete set of
file-specific metadata (Figure~\ref{fig1:FrameWork}).  Moreover, it is
a software implementation of the Gating-ML CR, an emerging new flow
cytometry data standard developed in collaboration with the ISAC Data
Standards Task Force, which makes it possibile to integrate
\Rpackage{flowCore} in existing work flows and to communicate with any
other FCM tool that adheres to the proposed standard. Adherence to
standards also plays a critical role in the ability of new methods
based on \Rpackage{flowCore} to find their way back into the standard
practices for FCM data analysis. \Rpackage{flowCore} is not a
GUI-driven software, and all operations are done using a command line
interface.  It is possible to add a more elaborate user interface on
top of this infrastructure, however the focus in this paper is on a
programmatic approach to enable the convenient development of novel
analysis methods and automation of complex analysis approaches.  By
taking the burden of data management from the programmer, and by
providing well-defined programming interfaces (APIs), it is possible
to readily test new ideas and to easily extend the framework's
functionality.

@@

In addition to the \Rpackage{flowCore} package that offers basic
infrastructure, we have implemented a range of additional Bioconductor
packages that are dedicated to more specific tasks of FCM data
analysis. The \Rpackage{flowViz} package \citep{sarkar2008ufv}
provides sophisticated data visualization tools, that make use of
multivariate trellis plotting \citep{lattice}.  These functions can be
used to quickly generate customized plots for extended cytometry data
sets for both direct data inspection and quality control.  The objects
metadata information can be used to arrange the layout and composition
of the plots.  Furthermore, the design and the API of the
visualization software is very generic, and users can readily extend
its capabilities by providing self-defined plotting functions.  The
\Rpackage{flowQ} package offers more advanced quality assurance (QA)
methodology and a framework to create interactive web-based reports of
QA results.  Most of \Rpackage{flowUtil}'s content deals with data
import and export, implementing the flow-cytometry specific standard
markup language.  Finally, \Rpackage{flowStats} provides more
elaborate statistical methods that are relevant in the context of flow
cytometry data analysis. \cite{lo2008agf} have recently developed an
automatic gating approach via robust model-based clustering using
\Rpackage{flowCore}'s data model and infrastructure which is
implemented in the Bioconductor package \Rpackage{flowClust} Another
package, \Rpackage{plateCore}, providing more specialized support for
experiments conducted on microtitre plates and facilitiating the
handling of spatial metadata, is under development.

>>>>

The notion of classes from an object-oriented programming language
provides one coherent way to describe these richer data structures,
and functions or methods that work on them allow for interaction and
manipulation. Many of the currently available software solutions offer
only limited support for such self-contained structures, or make use
of containers like XML or binary storage formats that are designed
specifically for particular user interfaces and hence are not easily
amenable to programmatical access. In addition, the closed-source
nature of these products often makes it impractical to integrate them
into analysis pipelines. In this manuscript we describe classes for
FCM data analysis and their implementation in R, however, they could
just as easily be implemented in any other language (e.g., Java,
C++). Software written in those languages could use similar data
structures, thereby simplifying communication and the interchange of
data between analysis tools.

The \Rpackage{flowCore} framework presented here can import and
process raw data FCS files along with their complete set of
file-specific metadata (Figure~\ref{fig1:FrameWork}).  Moreover, it is
a software implementation of the Gating-ML CR, an new flow
cytometry data standard developed in collaboration with the ISAC Data
Standards Task Force, which makes it possibile to integrate
\Rpackage{flowCore} in existing work flows and to communicate with any
other FCM tool that adheres to the proposed standard\citep{SpidlenInPressCytometryA}. Adherence to
standards also plays a critical role in the ability of new methods
based on \Rpackage{flowCore} to find their way back into the standard
practices for FCM data analysis. \Rpackage{flowCore} is not a
GUI-driven software, and all operations are done using a command line
interface.  It is possible to add a more elaborate user interface on
top of this infrastructure, however the focus in this paper is on a
programmatic approach to enable the convenient development of novel
analysis methods and automation of complex analysis approaches.  By
taking the burden of data management from the programmer, and by
providing well-defined programming interfaces (APIs), it is possible
to readily test new ideas and to easily extend the framework's
functionality.

@@

In addition to the \Rpackage{flowCore} package that offers basic
infrastructure, we have implemented a range of additional Bioconductor
packages that are dedicated to more specific tasks of FCM data
analysis. The \Rpackage{flowViz} package \citep{sarkar2008ufv}
provides sophisticated data visualization tools, that make use off
multivariate trellis plotting \citep{lattice}.  These functions can be
used to quickly generate customized plots for extended cytometry data
sets for both direct data inspection and quality control.  The objects
metadata information can be used to arrange the layout and composition
of the plots.  Furthermore, the design and the API of the
visualization software is very generic, and users can readily extend
its capabilities by providing self-defined plotting functions.  The
\Rpackage{flowQ} package offers more advanced quality assurance (QA)
methodology and a framework to create interactive web-based reports of
QA results.  Most of \Rpackage{flowUtil}'s content deals with data
import and export, implementing the flow-cytometry specific standard
markup language.  Finally, \Rpackage{flowStats} provides more
elaborate statistical methods that are relevant in the context of flow
cytometry data analysis. \cite{lo2008agf} have recently developed an
automatic gating approach via robust model-based clustering using
\Rpackage{flowCore}'s data model and infrastructure which is
implemented in the Bioconductor package \Rpackage{flowClust}. Another
package, \Rpackage{plateCore}, providing more specialized support for
experiments conducted on microtitre plates and facilitiating the
handling of spatial metadata, is under development.
